Current and future constraints on primordial magnetic fields

### 2016-11-03 pmf.tex
a

Measurements of the cosmic microwave background (CMB) temperature anisotropy have provided some of the most powerful tests of cosmology. 
We are now entering a new era as experiments begin to measure  polarized ``B-modes" in the CMB for the first time \citep{hanson13,polarbear14b,bicep2014a,naess14,keisler15,bicepkeck15,louis16}. 
Precision measurements of CMB polarization promise new tests of the standard cosmological model. 
The best known of these tests are the searches for inflationary gravitational waves in B-modes at large angular scales and plans to measure the sum of the neutrino masses through the lensing B-modes on small angular scales \citep[for a recent review see,][]{abazajian16}.

@@

The density and stress perturbations introduced by PMFs give rise to vorticity and gravitational waves in the early photon-baryon  plasma, i.e. vector and tensor modes. 
The exact amplitude of these modes can be calculated by modifying the normal Boltzmann equations to includes sources due to the PMFs. 
We use the CMBACT code\footnote{\url{http://www.sfu.ca/~levon/cmbact.html}} \citep{pogosian99}  to do this calculation. 
The PMF anisotropy is assumed to be drawn from a Gaussian distribution with a power law power spectrum, $A_b k^{n_b}$. 
Here $n_b$ is the spectral index of PMF and $A_b$ is its amplitude. 
We can translate the amplitude, $A_b$, into more physically-motivated units by calculating the rms of the PMF field strength over a length scale of 1\rm{Mpc}, thereby defining $B_{\rm{1Mpc}}$. 
Finally, the timing of when the PMFs are generated matters because the PMF-induced stress anisotropy can be compensated by decoupled or partially-decoupled neutrinos. 
The timing is typically parametrized by the ratio of the scale factors at neutrino decoupling and PMF generation, $a_{\nu}/a_{\rm{PMF}}$.

Given that we are still in the era of upper limits,  we do not explore the full 3-dimensional PMF parameter space in this work. 
Instead, we use CMBACT to calculate a PMF template for the CMB TT, TE, EE and BB power spectra at fixed parameters: $B_{\rm{1Mpc}}$=2.5 \rm{nG}, $n_b$=-2.9 and $a_{\nu}/a_{\rm{PMF}}=10^9$. 
This template is plotted against other signals in the CMB B-mode power spectrum in Fig.~\ref{fig:pmf-bb}. 
We fit for a new scaling parameter, $A_{pmf}$, that rescales the PMF power spectrum template. 
We can translate the amplitude constraint on this template to a constraint rms magnetic field strength on 1\,Mpc scales , \bpmf, using the expected scaling:
\be
\apmf \propto \bpmf^4.
\ee
An unfortunate consequence of this steep scaling is that substantial reductions in the observationally allowed power lead to only modestly better limits on the magnetic field strength. 
However, if a PMF is detected, this same steep scaling would allow excellent constraints on its properties.

@@

\begin{figure*}[htb]\centering
\includegraphics[width=0.9\textwidth,clip,trim={1.5cm 12.cm 5cm 4cm}]{pmf_templates.pdf}
  \caption[CMB polarization from PMFs]{
  The expected contributions to CMB B-modes. 
  The red line shows the expected lensing B-mode power in the \planck{} best-fit \lcdm{} cosmology. 
  The dark green line corresponds to the contribution from inflationary gravitational waves for $r=0.05$; the solid black line is the sum of the inflationary gravitational wave and lensing B-modes. 
  The dashed purple line shows the PMF template used in this work, scaled to \apmf{} = 0.2 to match the gravitational wave signal. 
  As can be seen, the tensor contribution from PMFs at large angular scales (low $\ell$) closely resembles the expected inflationary gravitational wave signal, and the two signals are effectively degenerate in a power spectrum (2-point) analysis.
  However, if excess power on these scales is detected, higher order moments of the map (i.e.~4-point estimators) might be used to discriminate between   the inflationary and PMF explanations. 
      \label{fig:pmf-bb}
  }
\end{figure*}

@@

We propagate these limits on the observed PMF power into limits on the magnetic field strength, \bpmf. 
Due to the scaling with the fourth power, an upper limit with a flat prior on \apmf{} would lead to an apparent `detection' of \bpmf. 
We therefore importance sample the chains to generate a flat prior on \bpmf. 
We find a 95\% limit of $\bpmf < 1.97$\,nG for \planck{} alone, completely consistent with \citep{planck15-19} who found $\bpmf < 2.0$\,nG with $n_B=-2.9$ held fixed as we've done. 
This is slightly reduced to $\bpmf < 1.85$\,nG for the combined dataset. 
The improvement from adding more data is somewhat less than one might predict because the combined likelihood peaks somewhat above zero, whereas the \planck{}-only constraints peaks at zero.

@@

The sensitivity of CMB experiments is increasing rapidly due to the exponential growth in detector counts. 
\citet{abazajian16} thus defines four stages of CMB experiments. 
The current experiments (i.e.~the ones used for PMF constraints in the last section) are classified as Stage II experiments. 
Stage III experiments such as SPT-3G, the Simons Array, or AdvACTPol \citep{benson14,suzuki16,henderson16} have approximately ten times more detectors, and generically will start collecting data in 2017 and finish in 2020-21.
In this section, we forecast the expected constraints from the Stage III experiments by combining forecasts for SPT-3G and the Simons Array. 
There is also a proposal to build a stage IV experiment, CMB-S4,  that would increase the detector counts by another order of magnitude and hopefully begin taking data in the early 2020s.
We examine the likely PMF constraints from CMB-S4, and consider which aspects of experimental or survey design are important to maximize the recovered information on PMFs.

@@

We use Fisher matrices to forecast the constraints on PMFs possible from each generation of experiment. 
While Fisher matrices may underestimate the true uncertainties for non-Gaussian likelihoods \citep[see e.g.,][]{wolz12}, they 
 make it easy to combine different experiments and estimate the final parameter constraints.

@@

We consider the constraints on PMFs in two cosmologies. 
The first is a 11-parameter model that extends \lcdm{} with 5 commonly considered extensions:  \lcdm{}  +r + \nrun{} + \neff{} + \mnu{}+ \apmf. 
This 11-parameter model is our default cosmological model for forecasting PMF constraints. 
Additionally, in order to illustrate the degree to which parameter degeneracies are limiting the inferred constraints,  we show constraints from a `minimal' 7-parameter model in which the PMF power is the only extension to \lcdm{}:  \lcdm{}  + \apmf. 
In both cases, we also marginalize over unknown Poisson EE and BB  terms due to polarized extragalactic sources.

In this work, we only forecast the uncertainties due to power spectrum (i.e.~2-point estimators) searches for PMFs. 
Currently the power spectrum limits from \planck{} or \pb{} are better than the 4-point upper limits. 
Although outside the scope of this work, it would be interesting to extend this analysis to 4-point estimators for several reasons. 
First, the 4-point limits should improve faster as the noise level falls. 
Second, in the case of a detection, the 4-point estimators might discriminate between inflationary gravitational waves and tensor PMFs. 
Finally, a lensing estimator might be used to `de-lense' the B-mode power spectrum, thereby suppressing the lensed B-mode signal and allowing better limits on PMFs. 
Such de-lensing techniques have long been proposed for inflationary gravitational wave searches \citep[e.g.][]{knox02,kesden02,seljak04a,simard15} and more recently been demonstrated on real \planck{} data \citep{larsen16}.

@@

The experiments that will begin taking data in 2017 will dramatically improve constraints on the PMF power. 
With the minimal cosmological model, the 1-sigma forecasts for \planck{}+\ho{} is $\sigma(\apmf)=0.47$ (this is consistent with, although slightly worse than, the upper limits from the real data).  
Adding the two stage III experiments (\sptnew{} and \simons{}) reduces the uncertainty nearly 400-fold in this 7-parameter model to $\sigma(\apmf)=0.0012$. 
Including all eleven parameters marginally weakens the \planck+\ho{} constraints on \apmf{} by 23\%, and the stage III constraint by a substantial factor of 2.5. 
Even within the 11-parameter model however,  the addition of stage III CMB experiments improves the \apmf{} uncertainty by a factor of $\sim$\,180 to  $\sigma(\apmf)=0.0031$. 
We can expect substantially tighter constraints on PMFs by the end of the decade.

@@

The primary motivation behind the proposed CMB stage IV experiment commonly called CMB-S4 is to search for inflationary gravitational waves and neutrino masses, however, the experiment should also enable extremely sensitive searches for PMFs. 
We begin by making forecasts for the fiducial CMB-S4 configuration as laid out in Table~\ref{tab:experiments}.

@@

We thank Srinivasan Raghunathan and Federico Bianchini for valuable feedback on the manuscript. 
C. Reichardt is the recipient of an Australian Research Council Future Fellowship (FT150100074), and also acknowledges support from the University of Melbourne. 
This research used resources of the National Energy Research Scientific Computing Center, which is supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231. 
We acknowledge the use of the Legacy Archive for Microwave Background Data Analysis (LAMBDA). Support for LAMBDA is provided by the NASA Office of Space Science.

>>>>

Measurements of the cosmic microwave background (CMB) temperature anisotropy have provided some of the most powerful tests of cosmology. 
We are now entering a new era as experiments begin to measure  polarized ``B-modes" in the CMB for the first time \citep{hanson13,polarbear14b,bicep14a,naess14,keisler15,bicepkeck15,louis16}. 
Precision measurements of CMB polarization promise new tests of the standard cosmological model. 
The best known of these tests are the searches for inflationary gravitational waves in B-modes at large angular scales and plans to measure the sum of the neutrino masses through the lensing B-modes on small angular scales \citep[for a recent review see,][]{abazajian16}.

@@

The density and stress perturbations introduced by PMFs give rise to vorticity and gravitational waves in the early photon-baryon  plasma, i.e. vector and tensor modes, as well as scalar modes. 
The exact amplitude of these modes can be calculated by modifying the normal Boltzmann equations to includes sources due to the PMFs. 
We use the CMBACT code\footnote{\url{http://www.sfu.ca/~levon/cmbact.html}} \citep{pogosian99}  to do this calculation. 
The PMF anisotropy is assumed to be drawn from a Gaussian distribution with a power law power spectrum, $A_b k^{n_b}$. 
Here $n_b$ is the spectral index of PMF and $A_b$ is its amplitude. 
We can translate the amplitude, $A_b$, into more physically-motivated units by calculating the rms of the PMF field strength over a length scale of 1\rm{Mpc}, thereby defining $B_{\rm{1Mpc}}$. 
Finally, the timing of when the PMFs are generated matters because the PMF-induced stress anisotropy can be compensated by decoupled or partially-decoupled neutrinos. 
The timing is typically parametrized by the ratio of the scale factors at neutrino decoupling and PMF generation, $a_{\nu}/a_{\rm{PMF}}$.

Given that we are still in the era of upper limits,  we do not explore the full 3-dimensional PMF parameter space in this work. 
Instead, we use CMBACT to calculate a PMF template for the CMB TT, TE, EE and BB power spectra at fixed parameters: $B_{\rm{1Mpc}}$=2.5 \rm{nG}, $n_b$=-2.9 and $a_{\nu}/a_{\rm{PMF}}=10^9$. 
This template includes tensor and vector PMF contributions; scalar contributions are not implemented in the code. 
We expect the missing scalar power to degrade the \planck-only results in this work.
However the missing scalar power should be  unimportant for the EE and BB spectra that are driving the tightest constraints. 
This template is plotted against other signals in the CMB B-mode power spectrum in Fig.~\ref{fig:pmf-bb}. 
We fit for a new scaling parameter, $A_{pmf}$, that rescales the PMF power spectrum template. 
For vector modes, we can translate the amplitude constraint on this template to a constraint rms magnetic field strength on 1\,Mpc scales , \bpmf, using the expected scaling:
\be
\apmf \propto \bpmf^4.
\ee
Note that the situation is more complicated for the tensor PMF modes where this power law scaling has not been demonstrated. 
An unfortunate consequence of this steep scaling is that substantial reductions in the observationally allowed vector-mode power lead to only modestly better limits on the magnetic field strength. 
However, if a PMF is detected, this same steep scaling would allow excellent constraints on its properties.

@@

\begin{figure*}[htb]\centering
\includegraphics[width=0.9\textwidth,clip,trim={1.5cm 12.cm 5cm 4cm}]{pmf_templates.pdf}
  \caption[CMB polarization from PMFs]{
  The expected contributions to CMB B-modes. 
  The red line shows the expected lensing B-mode power in the \planck{} best-fit \lcdm{} cosmology. 
  The dark green line corresponds to the contribution from inflationary gravitational waves for $r=0.05$; the solid black line is the sum of the inflationary gravitational wave and lensing B-modes. 
  The dashed purple line shows the PMF template used in this work, scaled to \apmf{} = 0.2 to match the gravitational wave signal. 
  As can be seen, the tensor contribution from PMFs at large angular scales (low $\ell$) closely resembles the expected inflationary gravitational wave signal, and the two signals are effectively degenerate in a power spectrum (2-point) analysis.
  %However, if excess power on these scales is detected, higher order moments of the map (i.e.~4-point estimators) might be used to discriminate between   the inflationary and PMF explanations. 
      \label{fig:pmf-bb}
  }
\end{figure*}

@@

We propagate these limits on the observed PMF power into limits on the magnetic field strength, \bpmf. 
We assume the scaling appropriate to the vector modes, i.e. $\apmf \propto \bpmf^4$,  applies to both vector and tensor modes. 
Given this assumed scaling, the observed upper limits with a flat prior on \apmf{} would lead to an apparent `detection' of \bpmf. 
We therefore importance sample the chains to generate a flat prior on \bpmf. 
We find a 95\% limit of $\bpmf < 2.0$\,nG for \planck{} alone, completely consistent with \citep{planck15-19} who found $\bpmf < 2.0$\,nG with $n_B=-2.9$ held fixed as we've done. 
This is only slightly reduced to $\bpmf < 1.9$\,nG for the combined dataset. 
The improvement from adding more data is somewhat less than one might predict because the combined likelihood peaks somewhat above zero, whereas the \planck{}-only constraints peaks at zero. 
We also stress that this should not be treated as a robust limit as the tensor PMF modes are crucial to the improvement, and the assumed scaling has not been tested for the tensor PMF. 
We quote it nonetheless to illustrate that with a steep power law scaling, the relative improvement to \bpmf{} will be slower than the improvement to \apmf{} as more data is added.

@@

The sensitivity of CMB experiments is increasing rapidly due to the exponential growth in detector counts. 
\citet{abazajian16} thus defines four stages of CMB experiments. 
The current experiments (i.e.~the ones used for PMF constraints in the last section) are classified as Stage II experiments. 
Stage III experiments such as SPT-3G, the Simons Array, or AdvACTPol \citep{benson14,suzuki16,henderson16} have approximately ten times more detectors, and generically will start collecting data in 2017 and finish in 2020-21.
In this section, we forecast the expected constraints from EE and BB power spectrum measurements  of the Stage III experiments by combining forecasts for SPT-3G and the Simons Array. 
There is also a proposal to build a stage IV experiment, CMB-S4,  that would increase the detector counts by another order of magnitude and hopefully begin taking data in the early 2020s.
We examine the likely PMF constraints from the EE and BB power spectrum measurements with CMB-S4, and consider which aspects of experimental or survey design are important to maximize the recovered information on PMFs.

@@

We use Fisher matrices to forecast the constraints on PMFs possible from each generation of experiment. 
The Fisher matrix, $\mathcal{F}_{ij}$, can be defined as:
\be
\mathcal{F}_{ij} = \frac{1}{2} \frac{\partial^2 \chi^2}{\partial p_i \partial p_j},
\ee
for parameters, p. 
The inverse of the Fisher matrix will be the covariance matrix for the parameters. 
While Fisher matrices underestimate the true uncertainties for non-Gaussian likelihoods \citep[see e.g.,][]{wolz12}, they 
 make it easy to combine different experiments and estimate the final parameter constraints.

@@

We consider the constraints on PMFs in two cosmologies. 
The first is a 11-parameter model that extends \lcdm{} with 5 commonly considered extensions:  \lcdm{}  +r + \nrun{} + \neff{} + \mnu{}+ \apmf. 
This 11-parameter model is our default cosmological model for forecasting PMF constraints. 
We have examined the degree to which \apmf{} is degenerate with the other 10 parameters -- the only strong degeneracy is with the tensor-to-scalar ratio, r. 
The reason for this degeneracy is illustrated in Fig.~\ref{fig:pmf-bb}, which shows that the tensor mode power due to inflationary gravitational waves and PMFs is nearly identical. 
To illustrate the degree to which parameter degeneracies limit the inferred constraints,  we also quote constraints from a `minimal' 7-parameter model in which the PMF power is the only extension to \lcdm{}:  \lcdm{}  + \apmf. 
In both cases, we also marginalize over unknown Poisson EE and BB  terms due to polarized extragalactic sources.

In this work, we only forecast the uncertainties due to power spectrum (i.e.~2-point estimators) searches for PMFs. 
Currently the power spectrum limits from \planck{} or \pb{} are better than the 4-point upper limits. 
Although outside the scope of this work, it would be interesting to extend this analysis to 4-point estimators for several reasons. 
First, the 4-point limits should improve faster as the noise level falls. 
Second, in the case of a detection, the 4-point estimators would almost certainly come into play to learn more about the vector modes of the PMFs. 
The detection (or non-detection) of the vector PMF signal in the 4-point estimators could also be used to argue for whether it is more likely that any observed tensor power is due to  inflationary gravitational waves or  PMFs. 
Finally, a lensing estimator might be used to `de-lense' the B-mode power spectrum, thereby suppressing the lensed B-mode signal and allowing better limits on PMFs. 
Such de-lensing techniques have long been proposed for inflationary gravitational wave searches \citep[e.g.][]{knox02,kesden02,seljak04a,simard15} and more recently been demonstrated on real \planck{} data \citep{larsen16}.

@@

The experiments that will begin taking data in 2017 will dramatically improve constraints on the PMF power. 
With the minimal cosmological model, the 1-sigma forecasts for \planck{}+\ho{} is $\sigma(\apmf)=0.47$ (this is consistent with, although slightly worse than, the upper limits from the real data).  
Adding the EE and BB bandpowers from the two stage III experiments (\sptnew{} and \simons{}) reduces the uncertainty nearly 400-fold in this 7-parameter model to $\sigma(\apmf)=0.0012$. 
Including all eleven parameters marginally weakens the \planck+\ho{} constraints on \apmf{} by 23\%, and the stage III constraint by a substantial factor of 2.5. 
Even within the 11-parameter model however,  the addition of stage III CMB experiments improves the \apmf{} uncertainty by a factor of $\sim$\,180 to  $\sigma(\apmf)=0.0031$. 
We can expect substantially tighter constraints on PMFs by the end of the decade.

@@

The primary motivation behind the proposed CMB stage IV experiment commonly called CMB-S4 is to search for inflationary gravitational waves and neutrino masses, however, the experiment should also enable extremely sensitive searches for PMFs. 
We look at the prospects for these searches using the polarization information (EE and BB power spectra) from CMB-S4. 
We begin by making forecasts for the fiducial CMB-S4 configuration as laid out in Table~\ref{tab:experiments}, and then consider how changing the experimental design would impact the expected uncertainties.

@@

We thank Srinivasan Raghunathan and Federico Bianchini for valuable feedback on the manuscript. 
C. Reichardt is the recipient of an Australian Research Council Future Fellowship (FT150100074), and also acknowledges support from the University of Melbourne. 
CF acknowledges support from NASA grants NASA NNX16AJ69G and NASA NNX16AF39G. 
This research used resources of the National Energy Research Scientific Computing Center, which is supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231. 
We acknowledge the use of the Legacy Archive for Microwave Background Data Analysis (LAMBDA). Support for LAMBDA is provided by the NASA Office of Space Science.
